The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-03-19 Paper/jaccardbaetica.tex
add new changes

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in Baetica province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. Baetica became important production and distribution centre during the Roman Empire. However, it remains under debate how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) producer centres by analysing different workshops in Baetica province and 2) two Roman provinces such as Germania and Britannia as a receptor centres. They will be used to detect a relation between the distribution of amphoric stamps and the economic structure in both centres. Here we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. Given this fact, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as Baetica (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered as the liquid gold since it was used in different aspect of the daily life as cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in a large number of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

@@

The high demand of Roman provinces stimulated by the good condition of the Baetica lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundreds of amphora workshops were made to support this high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this production was organized and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of Baetica province in the Roman organization. Indeed, archaeological evidence shows a highly specialized production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

@@

Three case study have been studied in order to analyse the relation between production centres (Baetica) and consumption centres (Britannia and Germania). In the case of Baetica province, we want to identify the role of the stamps in the organisation of the workshop. In Roman provinces, our aim is detecting groups of stamps concentrated in a area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could affect due to different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamp in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province.

@@

The amphora workshops were situated in different locations in Baetica province, along the river Guadalquivir and its tributary Genil in order to detect similarities between stamps from workshops and spatial distance (see Fig.\ref{workshop}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/workshop}
\caption{Distribution of the workshops}
\label{workshop}
\end{figure}

@@

We studied a dataset of 3798 stamps collected from different Dressel 20 amphora workshops in Baetica province. The stamp database was compiled by CEIPAC database \citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). However, approximately the 70 \% of stamps cannot be tested due to fragmentation or incomplete information. Consequently, we discard integrate the fragmented stamps in our dataset. We finally filter a total sample of 987 stamps comprised of 130 different stamps from 81 workshops.

@@

The consumption of olive oil in Britannia was residual until starting the Roman conquest \citep{funari_corpus_1996,
carreras_abastecimiento_2003}.

It is well known that this product was not frequently consumed by autonomous population. The absence of olive oil importation before the conquest can be reflected by the shortage of this product until the arrival of high percentage of Dressel 20 amphorae to Britannia \citep[ 1]{carreras_britannia_1998}. Additionally, the land of Britannia showed inadequate for the olive oil production due to low environmental conditions. This issue was necessary for creating an important export apparatus from Baetica to supply Roman legions.
In this moment, we detect an increase of the olive oil exportation concurring with the displacement of Roman legions during the military campaigns \citep[161]{monfort_britanniaen_1998}. This fact will be spatial incidence in sites close at Hadrian Wall.  
Olive oil production in Baetica would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns. 
%debate intercambios puntuales o redes comerciales de abastecimiento

The presence of Dressel 20 stamps in military camps in Britannia have been widely studied in Roman archaeology \citep{carreras_britannia_1998}. This fact also indicates a possible governmental organisation allocate to reforce and supply of olive oil in military camps, such as Germania (Remesal Rodrıguez, 1986). However, it is unknown the type of economical system when managing or control the redistribution of olive oil in different sites in Britannia but it seems that it could have been in the main cities \citep[45]{funari_economic_2005}.

Finally, The increase in olive oil exports would experience a progressive slowdown
from the third century A.D., coinciding with the change in market strategy in the
Empire. In that moment, a progressive dicrease of Dressel 20 is documented being gradually replace for Dressel 23 amphorae.

@@

We analysed a dataset of 2219 stamps from different centres in Britannia. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
The stamp database was also compiled by CEIPAC database. The database contained anomalies that were eliminated using the same criteria cited above. We finally selected the centres with more or equal than 5 stamps. As a result, we studied a total of 1765 stamps comprised of 968 different stamps from 46 centres.
The centres located in Britannia can be seen in Fig.\ref{britannia}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/britmap.pdf}
\caption{Distribution of centres in Britannia}
\label{britannia}
\end{figure}

@@

The Roman conquest in Germania dates from the end of the 1st century BC, during Caesar campaign and subsequently continued by Augusto. The advance of the Roman legions toward Gallia opened a way though the Atlantic Ocean to Germania \citep{remesal_annona_1986,
remesal_baetica_2002}.

It is thought that this route would have been initiated previously through 
terrestrial transport, although recent studies question this position giving an important role to the Atlantic route for both the conquests of Germania and those of Britannia \citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

As the case of Britain, the consumption of olive oil in Germania was not detected before the conquest.

The need to supply roman legions created a logistic network to supply to the army. In this moment, we detect the arrival of olive oil amphorae to Germania evidenced by the presence of Dressel 20 amphora in different sites close to limes.

The roman studies about the presence of Dressel 20 amphorae has not been the same repercusion than the rest of the province. This can be explained due to the lack of data \citep{horacio2010llegada}.

As a consequence, we do not the role that commerce agents in the participation of the distribution and exportation of olive oil in Germania or, on the contrary, the Roman army was the responsible to supply the product in the own province \citep[156]{remesal_germn_2010}.

@@

We analysed a dataset of 2052 stamps placed in Germania. All the data was also compiled by CEIPAC database. We finally selected a total of 1621 stamps consist of 850 different stamps from 46 sites (see Fig. \ref{germania}). As previously before, the centres were selected with more or equal than 5 stamps.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania5.pdf}
\caption{Distribution of the sites in Germania. Most sites were located in limes during the Roman Empire}
\label{germania}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendro}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different amphora workshops in Baetica area. Colours are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{dendro}
\end{figure}

@@

The results show a correlation similar to the obtained in Baetica province. However, the similarity is less pronounced as we observed in the dendrogram in Fig. \ref{britmap}

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendrobrit5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorthm of different sites in Britannia. Colours are represented by type of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{britmap}
\end{figure}

A minor correlation can be explained by the fact that Britannia spatial area is much wide than \textit{Baetica} where the centres were more concentrated.

Most military sites that showed greater similarity in the results were geographically close. We identified two factors in the results: 1) nearby sites tend to share similar stamps and 2) most sites have different stamps. 
By contrast, most sites did not show a strong similarity in stamp correlation. Neither we found a grouping of similar stamps in a specific place. In general, we did not observe any production pattern that indicates a clear organisation between production centres and reception centres. In other words, our results did not indicate the presence of a specific production center from Baetica in Britannia province. Rather it seems that olive oil production was distributed by non specific production centres.

@@

The similarity of stamps shows a results less significant than Britannia province as it can be seen in the dendrogram (Fig. \ref{germap}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different sites in Germania. Colours are represented by types of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{germap}
\end{figure}

@@

In this work, we aimed to analyse whether amphoric stamps could play an important role in the organization of the workshops along rivers. For this reason, dissimilarity index was used to detect differences among workshops and stamps.

@@

The hypothesis about groups of amphora workshops sharing the same stamps seems do not match with the results of the analysis even though there are similar stamps in closer workshops. Rather, it seems that each workshop were organized independently with different stamps. Those stamps detected in closer workshops do not move from other distant workshops. In other words, the stamps tend to keep in the same area and different stamps were located in a same amphora workshops.

This could be defined by several factors. First, each workshop had a different organization involved to the use of stamps and they were not used in other workshops. Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probably than closer workshops tend to share more traits than distant workshops. While the role of river was significant for the distribution of amphorae in consumption places, river connection amongst workshops does not seem show a relevancy for the distribution of stamps. Finally, the distribution of stamps could have shown some research bias. In some cases, workshops have been catalogued with different names in spite of belonging to the same workshops or being closer between each other. Additionally, most of the workshops were not widely excavated.

>>>>

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became important production and distribution centre during the Roman Empire. However, it remains under debate how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) producer centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as a receptor centres. They will be used to detect a relation between the distribution of amphoric stamps and the economic structure in both centres. Here we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. Given this fact, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as \textit{Baetica} (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered as the liquid gold since it was used in different aspect of the daily life as cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in a large number of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

@@

The high demand of Roman provinces stimulated by the good condition of the \textit{Baetica} lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundreds of amphora workshops were made to support this high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

%Hyphotesis
Nevertheless, some challenges remain under discussion such as how this production was organised and whether it is possible to distinguish production patterns in the olive oil trade. Our questions will be focused on the distribution of amphoric stamps. Did they follow a distribution pattern? Did stamps share the same workshop? Neither written records have been found that it could explain the economic role of \textit{Baetica} province in the Roman organisation. Indeed, archaeological evidence shows a highly specialised production with a long activity in a this specific area with apparently few changes \citep{remesal_anforas_2004}.

@@

Three case study have been studied in order to analyse the relation between production centres (\textit{Baetica}) and consumption centres (\textit{Britannia} and \textit{Germania}). In the case of \textit{Baetica} province, we want to identify the role of the stamps in the organisation of the workshop. In Roman provinces, our aim is detecting groups of stamps concentrated in a area or if some groups have an important role for the exportation of olive oil in those provinces. This economic connection could affect due to different aspects: a) correlation between spatial distance and centres based on the idea that closer workshops concentrate similar amphoric stamp in a specific area than the distant workshops and b) groups of similar stamps were concentrated in a specific Roman province.

@@

The amphora workshops were situated in different locations in \textit{Baetica} province, along the river Guadalquivir and its tributary Genil in order to detect similarities between stamps from workshops and spatial distance (see Fig.\ref{workshop}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/workshop1}
\caption{Distribution of the workshops}
\label{workshop}
\end{figure}

@@

We studied a dataset of 3798 stamps collected from different Dressel 20 amphora workshops in \textit{Baetica} province. The stamp database was compiled by CEIPAC database \citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). However, approximately the 70 \% of stamps cannot be tested due to fragmentation or incomplete information. Consequently, we discard integrate the fragmented stamps in our dataset. We finally filter a total sample of 987 stamps comprised of 130 different stamps from 81 workshops.

@@

The consumption of olive oil in \textit{Britannia} was residual until starting the Roman conquest \citep{funari_corpus_1996,
carreras_abastecimiento_2003}.

It is well known that this product was not frequently consumed by autonomous population. The absence of olive oil importation before the conquest can be reflected by the shortage of this product until the arrival of high percentage of Dressel 20 amphorae to \textit{Britannia} \citep[ 1]{carreras_britannia_1998}. Additionally, the land of \textit{Britannia} showed inadequate for the olive oil production due to low environmental conditions. This issue was necessary for creating an important export apparatus from \textit{Baetica} to supply Roman legions.
In this moment, we detect an increase of the olive oil exportation concurring with the displacement of Roman legions during the military campaigns \citep[161]{monfort_britanniaen_1998}. This fact will be spatial incidence in sites close at Hadrian Wall.  
Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns. 
%debate intercambios puntuales o redes comerciales de abastecimiento

The presence of Dressel 20 stamps in military camps in \textit{Britannia} have been widely studied in Roman archaeology \citep{carreras_britannia_1998}. This fact also indicates a possible governmental organisation allocate to reforce and supply of olive oil in military camps, such as \textit{Germania} (Remesal Rodrıguez, 1986). However, it is unknown the type of economical system when managing or control the redistribution of olive oil in different sites in \textit{Britannia} but it seems that it could have been in the main cities \citep[45]{funari_economic_2005}.

Finally, The increase in olive oil exports would experience a progressive slowdown
from the third century A.D., coinciding with the change in market strategy in the
Empire. In that moment, a progressive decrease of Dressel 20 is documented being gradually replace for Dressel 23 amphorae.

@@

We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
The stamp database was also compiled by CEIPAC database. The database contained anomalies that were eliminated using the same criteria cited above. We finally selected the centres with more or equal than 5 stamps. As a result, we studied a total of 1765 stamps comprised of 968 different stamps from 46 centres.
The centres located in \textit{Britannia} can be seen in Fig.\ref{britannia}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/britmap2.pdf}
\caption{Distribution of centres in \textit{Britannia}}
\label{britannia}
\end{figure}

@@

The Roman conquest in \textit{Germania} dates from the end of the 1st century BC, during Caesar campaign and subsequently continued by Augusto. The advance of the Roman legions toward \textit{Gallia} opened a way though the Atlantic Ocean to \textit{Germania} \citep{remesal_annona_1986,
remesal_baetica_2002}.

It is thought that this route would have been initiated previously through terrestrial transport, although recent studies question this position giving an important role to the Atlantic route for both the conquests of \textit{Germania} and those of \textit{Britannia} \citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

As the case of Britain, the consumption of olive oil in \textit{Germania} was not detected before the conquest.

The need to supply roman legions created a logistic network to supply to the army. In this moment, we detect the arrival of olive oil amphorae to \textit{Germania} evidenced by the presence of Dressel 20 amphora in different sites close to limes.

The roman studies about the presence of Dressel 20 amphorae has not been the same repercussion than the rest of the province. This can be explained due to the lack of data \citep{horacio2010llegada}.

As a consequence, we do not the role that commerce agents in the participation of the distribution and exportation of olive oil in \textit{Germania} or, on the contrary, the Roman army was the responsible to supply the product in the own province \citep[156]{remesal_germn_2010}.

@@

We analysed a dataset of 2052 stamps placed in \textit{Germania}. All the data was also compiled by CEIPAC database. We finally selected a total of 1621 stamps consist of 850 different stamps from 46 sites (see Fig. \ref{germania}). As previously before, the centres were selected with more or equal than 5 stamps.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania5_2.pdf}
\caption{Distribution of the sites in \textit{Germania}. Most sites were located in limes during the Roman Empire}
\label{germania}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendro}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different amphora workshops in \textit{Baetica} area. Colours are represented by areas divided into \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{dendro}
\end{figure}

@@

The results show a correlation similar to the obtained in \textit{Baetica} province. However, the similarity is less pronounced as we observed in the dendrogram in Fig. \ref{britmap}

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendrobrit5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different sites in \textit{Britannia}. Colours are represented by type of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{britmap}
\end{figure}

A minor correlation can be explained by the fact that \textit{Britannia} spatial area is much wide than \textit{Baetica} where the centres were more concentrated.

Most military sites that showed greater similarity in the results were geographically close. We identified two factors in the results: 1) nearby sites tend to share similar stamps and 2) most sites have different stamps. 
By contrast, most sites did not show a strong similarity in stamp correlation. Neither we found a grouping of similar stamps in a specific place. In general, we did not observe any production pattern that indicates a clear organisation between production centres and reception centres. In other words, our results did not indicate the presence of a specific production center from \textit{Baetica} in \textit{Britannia} province. Rather it seems that olive oil production was distributed by non specific production centres.

@@

The similarity of stamps shows a results less significant than \textit{Britannia} province as it can be seen in the dendrogram (Fig. \ref{germap}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different sites in \textit{Germania}. Colours are represented by types of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{germap}
\end{figure}

@@

In this work, we aimed to analyse whether amphoric stamps could play an important role in the organisation of the workshops along rivers. For this reason, dissimilarity index was used to detect differences among workshops and stamps.

@@

The hypothesis about groups of amphora workshops sharing the same stamps seems do not match with the results of the analysis even though there are similar stamps in closer workshops. Rather, it seems that each workshop were organised independently with different stamps. Those stamps detected in closer workshops do not move from other distant workshops. In other words, the stamps tend to keep in the same area and different stamps were located in a same amphora workshops.

This could be defined by several factors. First, each workshop had a different organisation involved to the use of stamps and they were not used in other workshops. Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probably than closer workshops tend to share more traits than distant workshops. While the role of river was significant for the distribution of amphorae in consumption places, river connection amongst workshops does not seem show a relevancy for the distribution of stamps. Finally, the distribution of stamps could have shown some research bias. In some cases, workshops have been catalogued with different names in spite of belonging to the same workshops or being closer between each other. Additionally, most of the workshops were not widely excavated.
